---
title: "first_steps"
author: "Githubz0r"
date: "2/13/2020"
output: html_document
---

## preparation
Load data and packages. We'll just both the data set with numCon/numEnds annotated by Silja and the one annotated by me.
```{r}
library(tidyverse)
library(irlba)
library(magrittr)
library(cowplot)
pancrea_holes_all <- read.csv('/Users/Deathvoodoo/Documents/semb_int/TableTilLars.csv')
pancrea_holes_altered <- read.csv('/Users/Deathvoodoo/Documents/semb_int/table_2_full.csv')
pancrea_holes_1 <- pancrea_holes_all[names(pancrea_holes_all)!='hole_area_raw'] %>% na.omit
pancrea_holes_2 <- pancrea_holes_altered[names(pancrea_holes_altered)!='hole_area_raw'] %>% na.omit
```

## analysis
First some plots
We just quickly visualize dAdt as a function of each predictor.
```{r}
predictors_1 <- c("numCon", "numEnds", "period", "hole_area_smooth", "eccentricity", "solidity")
predictors_2 <- c("numCon", "numEnds", "period", "hole_area_smooth", "eccentricity", "solidity") # same as predictors 1 but originally meant to include time
#predictors_2 <- c("numCon", "numEnds", "period", "hole_area_smooth", "eccentricity", "solidity", "time_point")
pancrea_1_norm <- pancrea_holes_1 %>% lapply(function(x){if (x[1] %>% is.numeric()) {x/sd(x)} else{x}}) %>% as.data.frame()
pancrea_2_norm <- pancrea_holes_2 %>% lapply(function(x){if (x[1] %>% is.numeric()) {x/sd(x)} else{x}}) %>% as.data.frame()

pancrea_1_norm_long <- pancrea_1_norm %>% pivot_longer(cols=predictors_1, names_to='predictors')
pancrea_2_norm_long <- pancrea_2_norm %>% pivot_longer(cols=predictors_2, names_to='predictors')

pancrea_1_norm_long %>% ggplot(aes(x=value, y=dAdt, col=predictors))+geom_point(size=0.2, alpha=0.5)
pancrea_2_norm_long %>% ggplot(aes(x=value, y=dAdt, col=predictors))+geom_point(size=0.2, alpha=0.5)
```

Now we do some PCA plots, with and without scaling the variances. We color both by hole and dataset to get an overview of how our points are lying together.
```{r}
DoPCA <- function(data.mat, predictor.names, make.plots=T, get.percent.variance=T, scale=T){
  panc.pca <- prcomp(data.mat[predictor.names], scale.=scale)
  if(make.plots){
    variances <- panc.pca$sdev^2
    PC.strings <- paste0('PC', '', 1:length(predictor.names))
    col.names <- paste0(PC.strings, ", % variance: ", 100*round(variances/sum(variances), digits = 4))
    panc.pca.df <- setNames(panc.pca$x %>% as.data.frame, col.names) %>% bind_cols(data.mat[setdiff(names(data.mat), predictor.names)])
    phole.scaled <- panc.pca.df %>% ggplot(aes(x=!!as.symbol(col.names[1]), y=!!as.symbol(col.names[2]), col=as.factor(hole_num_rep),
                                               shape=data_set_name))+geom_point(alpha=0.5)+guides(col = guide_legend(ncol = 3))
    pdataset.scaled <- panc.pca.df %>% ggplot(aes(x=!!as.symbol(col.names[1]), y=!!as.symbol(col.names[2]), col=data_set_name))+
      geom_point(alpha=0.5)+guides(col = guide_legend(ncol = 3))
  }
  return(list(panc.pca=panc.pca, phole.scaled=phole.scaled, pdataset.scaled=pdataset.scaled))
}

pancrea_1_pca <- DoPCA(pancrea_holes_1, predictors_1)
pancrea_2_pca <- DoPCA(pancrea_holes_2, predictors_2)

pancrea_1_pca$phole.scaled
pancrea_1_pca$pdataset.scaled

pancrea_2_pca$phole.scaled
pancrea_2_pca$pdataset.scaled

ggsave('/Users/Deathvoodoo/Documents/semb_int/pca_hole.png', pancrea_1_pca$phole.scaled)
ggsave('/Users/Deathvoodoo/Documents/semb_int/pca_dataset.png', pancrea_1_pca$pdataset.scaled)
```


Check weights of the features in the rotation matrices
```{r}
predictors_2 %>% setNames(1:length(predictors_2), .)
pancrea_1_pca$panc.pca$rotation
pancrea_2_pca$panc.pca$rotation
```


We check distribution of dAdt visually and with shapiro wilk test. Of course, for fitting a linear model (with gaussian noise) we assume that the data is normally distributed, so it better be.
```{r}
hist(pancrea_holes_1$dAdt, breaks = 50, prob=T)
lines(density(pancrea_holes_1$dAdt), col = "blue")
shapiro.test(pancrea_holes_1$dAdt)
```


Check correlation between variables, both as scatter plot and heat map (get different color scheme with two colors)
```{r}
MakeCorrelationPlots <- function(data.matrix){
  corr.mat <- cor(data.matrix)
  corr.values <- corr.mat; corr.values[upper.tri(corr.values, diag=T)] <- NA
  corr.values <- corr.values %>% reshape2::melt() %>% na.omit %>% transmute(which_factors=paste0(Var1,"+", Var2), correlation=value)
  pcor.scatter <- corr.values %>% ggplot(aes(x=which_factors, y=correlation))+geom_point()+theme(axis.text.x = element_text(angle = 90, hjust = 1))
  pcor.heat <- corr.mat %>% reshape2::melt() %>% ggplot(aes(x=Var1, y=Var2, fill=value))+geom_tile()+geom_text(aes(label = round(value, 2)))+
    scale_fill_gradient2(low = 'blue', high = 'red')+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  return(list(pcor.scatter=pcor.scatter, pcor.heat=pcor.heat))
}
corplot_1 <- MakeCorrelationPlots(pancrea_holes_1[c('dAdt', predictors_1)])
corplot_2 <- MakeCorrelationPlots(pancrea_holes_2[c('dAdt', predictors_2)])

corplot_1$pcor.scatter
corplot_1$pcor.heat
corplot_2$pcor.scatter
corplot_2$pcor.heat

#ggsave('/Users/Deathvoodoo/Documents/semb_int/correlation_matrix.png', corplot_1$pcor.heat)
```

How did the area of the holes change over time again (using my annotated time points) ?
```{r}
pancrea_area_df <- pancrea_holes_2[c('data_set_name', 'hole_num_rep', 'hole_num_inmovie', 'hole_area_smooth', 'time_point')]
area_time_1 <- pancrea_area_df %>% filter(data_set_name=='pos1') %>% ggplot(aes(x=time_point, y=hole_area_smooth, col=as.factor(hole_num_inmovie)))+geom_point(size=0.2, alpha=0.5)
area_time_2 <-pancrea_area_df %>% filter(data_set_name=='pos2') %>% ggplot(aes(x=time_point, y=hole_area_smooth, col=as.factor(hole_num_inmovie)))+geom_point(size=0.2, alpha=0.5)
area_time_3 <-pancrea_area_df %>% ggplot(aes(x=time_point, y=hole_area_smooth, col=as.factor(hole_num_inmovie)))+geom_point(size=0.2, alpha=0.5) + 
  facet_wrap(~data_set_name, scale='free') + guides(col = guide_legend(ncol = 2))
area_time_3
#ggsave('/Users/Deathvoodoo/Documents/semb_int/area_time1.png', area_time_1)
#ggsave('/Users/Deathvoodoo/Documents/semb_int/area_time2.png', area_time_2)
#ggsave('/Users/Deathvoodoo/Documents/semb_int/area_time.png', area_time_3, width = 10)
```

Also we can check the distributions of each predictor variable.
```{r}
pancrea_holes_1[predictors_1] %>% pivot_longer(predictors_1, names_to = 'predictors') %>% ggplot(aes(x=value))+geom_histogram()+facet_wrap(~predictors, scale='free')
```

We may have some reservations about period, let's try to check the average period of each hole as a function of the number of data points of the hole (as a substitute for life-time)
```{r}
mean_period <- pancrea_holes_1 %>% group_by(hole_num_rep) %>% summarize(period_mean=mean(period), nrow_holes=as.numeric(table(hole_num_rep)))
mean_period %>% ggplot(aes(x=as.factor(nrow_holes), y=period_mean, col=as.factor(hole_num_rep)))+geom_point()+theme(axis.text.x = element_text(angle = -90, hjust = 1))

#mean_dadt <- pancrea_holes_1 %>% group_by(hole_num_rep) %>% summarize(dadt_mean=mean(dAdt), nrow_holes=as.numeric(table(hole_num_rep)), area_mean=mean(hole_area_smooth))
#mean_dadt %>% ggplot(aes(x=area_mean, y=dadt_mean, col=as.factor(hole_num_rep)))+geom_point()+theme(axis.text.x = element_text(angle = -90, hjust = 1))
```

Now we start fitting some models. We first try all the variables. Note that we are omitting dEdt because Silja said it's not important and also numEnds_over_numCon since it's very highly correlation with numEnds
```{r}
cor(pancrea_holes_1$numEnds, pancrea_holes_1$numEnds_over_numCon)
```

The third model will be including the time, i.e. the time in the video annotated by me.
```{r}
linmodel_1 <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = pancrea_holes_1)
linmodel_2 <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = pancrea_holes_2)
linmodel_time <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity + time_point, data = pancrea_holes_2)
```

Diagnostic plots of the models. Things definitely do not look iid/evenly distribution noise/homoskedastic
```{r}
plot(linmodel_1)
plot(linmodel_2)
plot(linmodel_time)
```

We print some summaries of the models. This gives some various statistical info, such as e.g. weight estimates, p values when doing some kind of t-test of the weights (with null hypothesis that they are equal to zero). This gives similar values to doing a likelihood ratio test where the nested model lacks that coefficient. Note that the models are all complete trash, judging by the R^2 values.
```{r}
summary(linmodel_1)
summary(linmodel_2)
summary(linmodel_time)
```

Just testing linmodel_time vs the nested model linmodel_2 (and we can also do some aic). We see that time basically sucks. 
```{r}
anova(linmodel_2, linmodel_time)
AIC(linmodel_2) - AIC(linmodel_time) # should be positive if linmodel 3 is better
```

Is our model even better than just random shit? We can make some random data frames and see.
```{r}
RandomizeDataframe <- function(df){
  df.random <- df %>% lapply(function(x){sample(x)}) %>% as.data.frame
  return(df.random)
}

random_pancreas <- replicate(100, RandomizeDataframe(pancrea_holes_1), simplify = FALSE)
```

Let's test the random data frames
```{r}
random_results <- random_pancreas %>% lapply(function(x){
  linmodel.random <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = x)
  random.lm.summary <- summary(linmodel.random)
  residual.standard.error <- random.lm.summary$sigma
  r.squared <- random.lm.summary$r.squared
  r.squared.adjusted <- random.lm.summary$adj.r.squared
  sum.squared.resids <- sum(residuals(linmodel.random)^2)
  return(c(residual.standard.error=residual.standard.error, r.squared=r.squared, 
           r.squared.adjusted=r.squared.adjusted, sum.squared.resids=sum.squared.resids))
})

output_values_random <-random_results[[1]] %>% names %>% lapply(function(x){
  value.for.all.res <- random_results %>% lapply(function(result.vector){result.vector[[x]] %>% unname})
  return(value.for.all.res)
}) %>% lapply(unlist); output_values_random %<>% setNames(random_results[[1]] %>% names)

output_values_random$residual.standard.error %>% qplot
output_values_random$r.squared %>% qplot
output_values_random$r.squared.adjusted %>% qplot
output_values_random$sum.squared.resids %>% qplot
```

We compare mean residual standard error, mean r squared, mean r squared adjusted and the mean sum of squared residuals from randomized data frames with the values for the first linear model. It's a little bit better.
```{r}
output_values_random %>% lapply(mean) %>% unlist
summary(linmodel_1)
sum(residuals(linmodel_1)^2)
```

Now we can do some stepwise linear regression with different R packages, using AIC/BIC as criteria.
```{r}
library(MASS)
library(leaps)
```

First using MASS
```{r}
step_mass_bi <- stepAIC(linmodel_1, direction='both')
summary(step_mass_bi)
```
We can try putting in a model with second order interactions and see what happens.
```{r}
step_mass_squared_bi <- stepAIC(lm(dAdt~.^2, data = pancrea_holes_1[c('dAdt', predictors_1)]), direction='both')
summary(step_mass_squared_bi)
```
Gives better adjusted R squared fo sho.


Using leaps package
```{r}
models_leap <- regsubsets(dAdt~., data = pancrea_holes_1[c('dAdt', predictors_1)], nvmax = 7,
                     method = "seqrep")

summary_leap <- summary(models_leap)
summary_leap
summary_leap$bic
```

Doing both forwards and backwards just because.
```{r}
models_forward <- regsubsets(dAdt~., data = pancrea_holes_1[c('dAdt', predictors_1)], nvmax = 7,
                     method = "forward")

summary_forward <- summary(models_forward)
summary_forward
summary_forward$bic
```

```{r}
models_backward <- regsubsets(dAdt~., data = pancrea_holes_1[c('dAdt', predictors_1)], nvmax = 7,
                     method = "backward")

summary_backward <- summary(models_backward)
summary_backward
summary_backward$bic
```

We can try higher order model, see wut happens.
```{r}
Nsquaredpermutations <- function(n.features){ # since the regsubsets function has a max number of variables, we just put in the maximum number of possible variables of 2nd order
  return((n.features^2-n.features)/2 + n.features + 1)
}
models_leap_square <- regsubsets(dAdt~.^2, data = pancrea_holes_1[c('dAdt', predictors_1)], nvmax = Nsquaredpermutations(length(predictors_1)),
                     method = "seqrep")

summary_leap_square <- summary(models_leap_square)
summary_leap_square
summary_leap_square$bic
```

I.e. the interaction model with the lowest bic is 
```{r}
which(summary_leap_square$bic==min(summary_leap_square$bic))
```
We extract the variables of this model and put it into lm
```{r}
interaction_str <- summary_leap_square$outmat[12, ][summary_leap_square$outmat[12, ]=='*'] %>% names %>% paste(., collapse='+') %>% paste0('dAdt~','',.)
linmodel_interaction_leaps <- lm(as.formula(interaction_str), data=pancrea_holes_1)
```

```{r}
summary(linmodel_interaction_leaps)
AIC(linmodel_interaction_leaps)
BIC(linmodel_interaction_leaps)
plot(linmodel_interaction_leaps)
```
Yeah, improved adjusted R squared fo sho.

Btw, which rows were removed by na.omit?
```{r}
removed_rows_bool <- pancrea_holes_all[names(pancrea_holes_all)!='hole_area_raw'] %>% apply(1, function(row){
  return((row %>% sapply(function(x){is.na(x)}) %>% sum())>0)
})
removed_rows <- pancrea_holes_all[removed_rows_bool, ]
table(removed_rows$hole_num_rep)
table(pancrea_holes_all$hole_num_rep) # 21 and 23 are completely removed by the filtering due to NaNs in period
```


Check similarity between me and siljas counts, using zero-one loss and squared loss.
```{r}
Zeroone <- function(v1, v2){
  return(1-sum(v1==v2)/length(v1))
}

Squarederror <- function(v1, v2){
  sq.loss <- sum((v1-v2)^2)/length(v1)
  return(sq.loss)
}

Squarederror(split(pancrea_holes_1$numEnds, pancrea_holes_1$hole_num_rep)$`22`, split(pancrea_holes_2$numEnds, pancrea_holes_2$hole_num_rep)$`22`) # sanity check

CalcError <- function(df1, df2, vectors.to.split, split.by='hole_num_rep'){
  calcerrors <- function(v1, v2){
    return(c(zeroone=Zeroone(v1, v2), squarederror=Squarederror(v1, v2)))
  }
  sub.dfs.by.factor <- vectors.to.split %>% lapply(function(vector.name){
    df1.split <- df1[[vector.name]] %>% split(df1[[split.by]])
    df2.split <- df2[[vector.name]] %>% split(df2[[split.by]])
    return(list(df1.split = df1.split, df2.split = df2.split))
  })
  sub.dfs.by.factor %<>% setNames(vectors.to.split)
  errors.by.factor <- sub.dfs.by.factor %>% lapply(function(data.list){
    errors.by.split <- Map(calcerrors, data.list[[1]], data.list[[2]])
    return(errors.by.split)
  })
  errors.by.factor <- errors.by.factor %>% lapply(function(x){do.call(rbind,x)}) %>% lapply(as.data.frame) %>% lapply(tibble::rownames_to_column, var='hole')
  errors.df <- dplyr::bind_rows(errors.by.factor, .id='compared_factor')
  average.errors <- vectors.to.split %>% sapply(function(vector.name){
    return(calcerrors(df1[[vector.name]], df2[[vector.name]]))
  })
  return(list(errors.df=errors.df, average.errors=average.errors))
}
```

```{r}
error_pancrea <- CalcError(pancrea_holes_1, pancrea_holes_2, c('numCon', 'numEnds'))
error_pancrea$errors.df %>% ggplot((aes(x=as.factor(as.numeric(hole)), y=zeroone, col=compared_factor)))+
  theme(axis.text.x = element_text(angle = -90, hjust = 1))+geom_jitter()
error_pancrea$errors.df %>% ggplot((aes(x=as.factor(as.numeric(hole)), y=squarederror, col=compared_factor)))+
  theme(axis.text.x = element_text(angle = -90, hjust = 1))+geom_jitter()
```

quick plot of fitted vs real values for the various models
```{r}
plot(linmodel_1$fitted.values~pancrea_holes_1$dAdt)
plot(step_mass_squared_bi$fitted.values~pancrea_holes_1$dAdt)
plot(linmodel_interaction_leaps$fitted.values~pancrea_holes_1$dAdt)
```

Let's also just check the correlations
```{r}
cor(linmodel_1$fitted.values, pancrea_holes_1$dAdt)
cor(step_mass_squared_bi$fitted.values, pancrea_holes_1$dAdt)
cor(linmodel_interaction_leaps$fitted.values, pancrea_holes_1$dAdt)
```



We can also try to down sample to circumvent IID assumption violation
```{r}
Sampleintervals <- function(df, interval.length, offset=0){
  start.from <- 1+offset
  n <- nrow(df)
  if(start.from >= n || interval.length >= n){
    indices <- sample(n, 1)
  } else {
    indices <- seq(start.from, n, interval.length)
  }
  return(df[indices, ])
}

Makedownsampled <- function(df, interval.length, split.by='hole_num_rep', offset=0){
  sub.dfs <- df %>% split(df[[split.by]])
  downsampled.dfs <- sub.dfs %>% lapply(function(x){return(Sampleintervals(x, interval.length=interval.length, offset=offset))})
  downsampled.rbound <- dplyr::bind_rows(downsampled.dfs)
  return(downsampled.rbound)
}

downsampled_df <- Makedownsampled(pancrea_holes_1, 5, 'hole_num_rep')
linmodel_downsampled <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = downsampled_df)
```

```{r}
plot(linmodel_downsampled)
```

```{r}
summary(linmodel_downsampled)
```
Things look less fucked, but we have less statistical power now. Quickly check pred vs true plot and correlation. The results are less nighmarish looking but clearly linear model is insufficient.
```{r}
plot(linmodel_downsampled$fitted.values~downsampled_df$dAdt)
cor(linmodel_downsampled$fitted.values, downsampled_df$dAdt)
```


We can try different offsets during the sampling (sampling 1 random point from the hole if offset >= n_hole)
```{r}
offsets_ <- c(0, 5, 10, 20, 30, 50)
dfs_offsets <- offsets_ %>% lapply(function(x, interval.length=20){
  df.downsampled <- Makedownsampled(pancrea_holes_1, interval.length, offset=x)
  return(df.downsampled)
})

linmodels_offsets <- dfs_offsets %>% lapply(function(x){
  lm.downsampled <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = x)
  return(lm.downsampled)
})

```

```{r}
offset_residplots <- linmodels_offsets %>% lapply(function(x){
  return(plot(x, which=c(1,1)))
})
```

summaries of the models
```{r}
linmodels_offsets %>% lapply(summary)
```

just a quick sanity test because I forgot why permuting doesn't change linear regression results, but ofc it's because permutation matrix cancels out in the formula
```{r}
quicktest <- lm(dAdt~numCon + numEnds + period + hole_area_smooth + eccentricity + solidity, data = pancrea_holes_1[sample(1:nrow(pancrea_holes_1)), ])
```



